---
title: "tutorial01-tidymodels"
author: "Nicholas R. Record"
date: 
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This markdown document steps through the "getting started" tuturial from Tidymodels <https://www.tidymodels.org/start/models/>.

You can click through the code to run it in real time, or  click the **Knit** button to generate an HTML that includes both content as well as the output of any embedded R code chunks within the document. 

The first chunk of code below loads the necessary libraries. If you don't have these libraries yet, you'll need to install them using the install.packages() command.

```{r}
library(tidymodels)  # for the parsnip package, along with the rest of tidymodels

# Helper packages
library(readr)       # for importing data
library(broom.mixed) # for converting bayesian models to tidy tibbles
library(dotwhisker)  # for visualizing regression results
```

## Loading the Sea Urchins Data

This code pulls the example data from a website

```{r}
urchins <-
  # Data were assembled for a tutorial 
  # at https://www.flutterbys.com.au/stats/tut/tut7.5a.html
  read_csv("https://tidymodels.org/start/models/urchins.csv") %>% 
  # Change the names to be a little more verbose
  setNames(c("food_regime", "initial_volume", "width")) %>% 
  # Factors are very helpful for modeling, so we convert one column
  mutate(food_regime = factor(food_regime, levels = c("Initial", "Low", "High")))
```

What does the data look like?
```{r}
urchins
```

And here is a fancy plot of the data. Don't worry if all of this code doesn't make sense yet.
```{r}
ggplot(urchins,
       aes(x = initial_volume, 
           y = width, 
           group = food_regime, 
           col = food_regime)) + 
  geom_point() + 
  geom_smooth(method = lm, se = FALSE) +
  scale_color_viridis_d(option = "plasma", end = .7)
```

## Build and fit a model

From here, you can add chunks of code from the tutorial page <https://www.tidymodels.org/start/models/>. Make sure the chunks of code are pasted within the R blocks, like below.

```{r}

```